Memories of migrations past: Sociality and cognition in dynamic, seasonal environments

### 2021-07-15 text/frontiers.tex
msmsmsmsmsmsms

\usepackage{url,lineno,microtype,subcaption}

\usepackage[onehalfspacing]{setspace}

\usepackage[colorlinks]{hyperref}
\hypersetup{citecolor={blue}}
\hypersetup{urlcolor={blue}}
\hypersetup{linkcolor={blue}}

@@

\def\keyFont{\fontsize{8}{11}\helveticabold }
\def\firstAuthorLast{Sample {et~al.}} %use et al only if is more than 1 author
\def\Authors{Eliezer Gurarie\,$^{1,2,*}$, 
	Sriya Potluri\,$^1$, 
	G.~Christopher Cosner\,$^3$,
	R.~Stephen Cantrell\,$^3$,
	William F.~Fagan\,$^1$
}
% Affiliations should be keyed to the author's name with superscript numbers and be listed as follows: Laboratory, Institute, Department, Organization, City, State abbreviation (USA, Canada, Australia), and Country (without detailed address information such as city zip codes or street names).
% If one of the authors has a change of address, list the new address below the correspondence details using a superscript symbol and use the same symbol to indicate the author in the author list.
\def\Address{$^{1}$Department of Biology, University of Maryland, College Park, MD 20742, USA.\\
$^{2}$Department of Environmental and Forest Biology, Syracyse, NY   \\
$^{3}$Department of Mathematics, The University of Miami, Coral Gables, FL 33146, USA.
}
% The Corresponding Author should be marked with an asterisk
% Provide the exact contact address (this time including street name and city zip code) and email of the corresponding author
\def\corrAuthor{Eliezer Gurarie}

\def\corrEmail{egurarie@umd.edu}

@@

Seasonal migrations are widespread among terrestrial, aquatic, avian and invertebrate species \citep{Dingle2014}. For many species, migration is an extremely successful strategy, allowing a far greater number of individuals to inhabit landscapes which might not otherwise be able to support large numbers year round \citep{Fryxell1988}. Essentially, the evolutionary stability of a migratory strategy relies on the fitness benefits of accessing suitable seasonal resources, whether those are for energetic gain, predator avoidance, or a suitable physical, biotic or social environment for reproduction outweighting the energetic and survival related costs of migration \citep{Avgar2014}.

Proximate causes, drivers and mechanism can vary considerably across and even within species \citep{Berthold1999, Shaw2016}. Some migrants follow a ``green wave'' of spring vegetation as it flowers across altitudinal or latitudinal gradients \citep{Bischof2012, Kolzsch2015}. These migrations can be considered ``tactical'', as they can occur - as an extreme simplification - purely as response to local conditions. Other migrants perform long-distance migrations in anticipation that critical resources will be available at the time of arrival at the end point of migration \citep{Abrahms2019}. This second behavior involves the greatest trade-off between the costs of migration against the benefits of accessing resources, whether those are food, suitable habitat for breeding, or refuge from predators, that are highly seasonal and localized. This approach can be considered ``strategic'' in the sense that it is driven not by immediate cues but by an anticipation based on prior experience \citep{Bracis2017}.

Migratory species are often considered to be more vulnerable to environmental change, as a disruption in either of the seasonal ranges or along a migratory corridor can have significant negative impacts \citep{Wilcove2008, Kauffman2021}. On the other hand, it has been argued that migratory species might be more resilient to disruptions due to their wide-ranging mobility \citep{Robinson2009}. Clearly, the ability to perform a migration without local cues is only possible if the behavior is hard-programmed or remembered. On the other hand, a strictly programmed behavior can be maladaptive if conditions change. Because the scenarios underlying migration are manifold and complex, mathematical modeling may provide some insights and help clarify where, when, and under what conditions we might migration behavior to emerge, to be adaptive, to be maladaptive, or to collapse.

Diffusion-advection models have a long pedigree in animal movement modeling \citep{Skellam1951, Turchin1998, Okubo2001}. These models are grounded in the general idea that animal movements, somewhat like movements of physical particles, combine random (diffusive) components with directed (advective) components. While direct relationships between diffusion models and movement data are somewhat tenuous \citep{Gurarie2011,Potts2020}, as a theoretical tool for exploring processes they are invaluable for their versatility and the relative ease of numeric computation of the partial differential equations (PDEs) that describe them mathematically. Much theoretical and some applied work has been done on refining the basic assumptions of diffusion models, e.g.~by including heterogeneity in populations \citep{Skalski2003, Gurarie2009}, fat-tailed dispersal kernels \citep{Kot1996}, non-linear or otherwise complex responses to resources and conspecifics. But perhaps the greatest difference between animals and randomly moving passive particles described by diffusion models is cognition, including social behavior and memory. Refinements to diffusion-advection equations have revealed conditions under which non-local information gathering \citep{Fagan2017} and behavioral switching may confer foraging advantages \citep{Fagan2019}, in particular when resources are dynamic and patchy. While diffusion models that incorporate social movements (swarming, schooling, herds) have been well-studied \citep{Okubo1986,Gruenbaum1994,Mogilner1999}, the interacting role of memory and sociality in a specifically migratory framework has been relatively unexplored.

Here, we develop a social diffusion-advection model with memory to explore the resilience of a migratory population under various dynamic, seasonal resource distributions. In formulating the model, our goal was to identify the minimum set of movement and memory parameters required to generate an adaptive, migratory behavior. This includes the ability to learn to migrate from non-migratory initial conditions, simulating the release of naive animals in a seasonal environment \citep{Jesmer2018}, to lose the propensity to migrate if the resource distribution does not require it, also a commonly observed phenomenon \citep{Wilcove2008}, and to assess the resilience or fragility of a migratory population against changing resource distribution dynamics, including both stochasticity and trends in spatial and temporal distributions, mirroring the effects of climate change.

We anticipated that under many conditions a blending of \emph{tactical} (i.e.~direct response to resource availability or perception) and \emph{strategic} (i.e.~memory-driven and forward-thinking) behavior will help foragers navigate dynamic, seasonal environments. Over-reliance on either strategy should be maladaptive. We further anticipate that a shorter-term memory updating is needed to navigate trends in resource spatial distribution and temporal distribution (phenology), but that a longer-term reference memory is needed to navigate resource distributions that are stochastic \citep{Lin2021}.  Similarly, we anticipated that a balance between very low sociality and extreme sociality would lead to the most resilient migratory process.

@@

Within a given year, the resource is entirely symmetric: $h_y(x,t) = h_y(-x, \tau-t)$. However, in scenarios exploring climate change we allow the peaks to vary with drift and stochasticity according to: $m_x(y) \sim {N}(\mu_x + \gamma_x\,y, \sigma_x)$ and $m_t(y) \sim {N}(\mu_t + \gamma_t\,y, \sigma_t)$, where the $\mu$, $\gamma$ and $\sigma$ terms are the mean, slope and variance, respectively, for the location and time duration of the pulse. Thus, if $\gamma=0$ and $\sigma=0$, the conditions are constant across years and if $\gamma_x > 0$ there is a drift of the resource towards extremes, if $\beta_t < 0$ there is a shift towards earlier resource pulses. These trends model the pole-ward shift of peak resources and the earlier spring phenology occurring with a warming global climate. The spatial scales ($s_x$ and $s_t$) remain constant in all of our simulations.

@@

All movement model parameters, resource parameters, and metrics are summarized in table \ref{table_definitions}.

@@

@@

\section*{Conflict of Interest Statement}

@@

EG and WFF provided the original idea. EG and SP developed and ran models and analysis. All authors contributed to the writing.

\section*{Funding}

EG, WFF, GCC and RSC, were supported in part by NSF award DMS 1853478. EG and WFF were further partially supported by NSF grant IIBR 1915347.

\section*{Acknowledgments}
The authors are grateful to Quentin Read at SESYNC for computational support and advice for multi-node multi-core model runs.

\section*{Data Availability Statement}

The only data analyzed in this manuscript was generated by simulation scripts made available in the GitHub repository at \url{https://github.com/EliGurarie/memorymigration}.

@@

>>>>

\usepackage{url,hyperref,lineno,microtype,subcaption}
%\usepackage[onehalfspacing]{setspace}

@@

\def\keyFont{\fontsize{8}{11}\helveticabold }
\def\firstAuthorLast{Sample {et~al.}} %use et al only if is more than 1 author
\def\Authors{Eliezer Gurarie\,$^{1,2}$, 
	Sriya Potluri\,$^1$, 
	G.~Christopher Cosner\,$^3$,
	R.~Stephen Cantrell\,$^3$,
	William F.~Fagan\,$^1$
}
% Affiliations should be keyed to the author's name with superscript numbers and be listed as follows: Laboratory, Institute, Department, Organization, City, State abbreviation (USA, Canada, Australia), and Country (without detailed address information such as city zip codes or street names).
% If one of the authors has a change of address, list the new address below the correspondence details using a superscript symbol and use the same symbol to indicate the author in the author list.
\def\Address{$^{1}$Department of Biology, University of Maryland, College Park, MD 20742, USA.\\
$^{2}$Department of Environmental and Forest Biology, Syracyse, NY   \\
$^{3}$Department of Mathematics, The University of Miami, Coral Gables, FL 33146, USA.
}
% The Corresponding Author should be marked with an asterisk
% Provide the exact contact address (this time including street name and city zip code) and email of the corresponding author
\def\corrAuthor{Corresponding Author}

\def\corrEmail{email@uni.edu}

@@

Seasonal migrations are extremely widespread among terrestrial, aquatic, avian and invertebrate species \citep{Dingle2014}. For many species, migration is an extremely successful strategy, allowing a far greater number of individuals to inhabit landscapes which might not otherwise be able to support large numbers year round \citep{Fryxell1988}. The ultimate evolutionary cause of seasonal migration, essentially, is that the energetic and survival related costs of migration are outweighed by the fitness benefits of accessing suitable seasonal resources, whether those are for energetic gain, predator avoidance, or a suitable physical, biotic or social environment for reproduction \citep{Avgar2014}.

Much as the ultimate causes of migration vary across species, the proximate causes, drivers and mechanism can vary considerably across and even within species \citep{Berthold1999, Shaw2016}. Some migrants follow a ``green wave'' of spring vegetation as it flowers across altitudinal or latitudinal gradients \citep{Bischof2012, Kolzsch2015}. These migrations can be considered ``tactical'', as they can occur - as an extreme simplification - purely as response to local conditions. Other migrants perform long-distance migrations in anticipation that critical resources will be available at the time of arrival at the end point of migration \citep{Abrahms2019}. This second behavior involves the greatest trade-off between the costs of migration against the benefits of accessing resources, whether those are food, suitable habitat for breeding, or refuge from predators, that are highly seasonal and localized. This approach can be considered ``strategic'' in the sense that it is driven not by immediate cues but by an anticipation based on prior experience \citep{Bracis2017}.

Migratory species are often considered to be more vulnerable to environmental change, as a disruption in either of the seasonal ranges or along a migratory corridor can have significant negative impacts \citep{Wilcove2008, Kauffman2021}. On the other hand, it has been argued that migratory species might be more resilient to disruptions due to their wide-ranging mobility \citep{Robinson2009}. Clearly, the ability to perform a migration without local cues is only possible if the behavior is hard-programmed or remembered. On the other hand, a strictly programmed behavior can be maladaptive if conditions change. Because the scenarios underlying migration are manifold and complex, mathematical modeling may provide some insights and help clarify where, when, and under what conditions we might expect different kinds of migrations to operate.

Diffusion-advection models have a long pedigree in animal movement modeling \citep{Skellam1951, Turchin1998, Okubo2001}. These models are grounded in the general idea that animal movements, somewhat like movements of physical particles, combine random (diffusive) components with directed (advective) components. While direct relationships between diffusion models and movement data are somewhat tenuous \citep{Gurarie2011,Potts2020}, as a theoretical tool for exploring processes they are invaluable for their versatility and the relative ease of numeric computation of the partial differential equations (PDEs) that describe them mathematically. Much theoretical and some applied work has been done on refining the basic assumptions of diffusion models, e.g.~by including heterogeneity in populations \citep{Skalski2003, Gurarie2009}, fat-tailed dispersal kernels \citep{Kot1996}, non-linear or otherwise complex responses to resources and conspecifics. But perhaps the greatest difference between animals and randomly moving passive particles described by diffusion models is cognition, including social behavior and memory. Refinements to diffusion-advection equations have revealed conditions under which non-local information gathering \citep{Fagan2017} and behavioral switching may confer foraging advantages \citep{Fagan2019}, in particular when resources are dynamic and patchy. While diffusion models that incorporate social movements (swarming, schooling, herds) have been well-studied \citep{Okubo1986,Gruenbaum1994,Mogilner1999}, the interacting role of memory and sociality is relatively unexplored.

Here, we develop a social diffusion-advection model with memory to explore the resilience of a migratory population under various dynamic, seasonal resource distributions. In formulating the model, our goal is to identify the minimum set of movement and memory parameters required to generate an adaptive, migratory behavior. This includes the ability to learn to migrate from non-migratory initial conditions, simulating the release of naive animals in a seasonal environment \citep{Jesmer2018}, to lose the propensity to migrate if the resource distribution does not require it, also a commonly observed phenomenon \citep{Wilcove2008}, and to assess the resilience (or fragility) of a migratory population with memory against changing resource distribution dynamics, including both stochasticity and trends in spatial and temporal distributions.

We anticipate that under many conditions a blending of \emph{tactical} (i.e.~direct response to resource availability or perception) and \emph{strategic} (i.e.~memory-driven and forward-thinking) behavior will help foragers navigate dynamic, seasonal environments. Over-reliance on either strategy should be maladaptive. We further anticipate that a shorter-term memory updating is needed to navigate trends in resource spatial distribution and temporal distribution (phenology), but that a longer-term reference memory is needed to navigate resource distributions that are stochastic \citep{Lin2021}.

@@

Within a given year, the resource is entirely symmetric: $h_y(x,t) = h_y(-x, \tau-t)$. However, in scenarios exploring climate change we allow the peaks to vary with drift and stochasticity according to: $m_x(y) \sim {N}(\mu_x + \gamma_x\,y, \sigma_x)$ and $m_t(y) \sim {N}(\mu_t + \gamma_t\,y, \sigma_t)$, where the $\mu$, $\gamma$ and $\sigma$ terms are the mean, slope and variance, respectively, for the location and time duration of the pulse. Thus, if $\gamma=0$ and $\sigma=0$, the conditions are constant across years and if $\beta_x > 0$ there is a drift of the resource towards extremes, if $\beta_t < 0$ there is a shift towards earlier resource pulses. These trends model the pole-ward shift of peak resources and the earlier spring phenology occurring with a warming global climate. The spatial scales ($s_x$ and $s_t$) remain constant in all of our simulations.

@@

All movement model parameters, resource parameters, and metrics are summarized in table \label{table_definitions}.

@@

\subsection{Figures}
Frontiers requires figures to be submitted individually, in the same order as they are referred to in the manuscript. Figures will then be automatically embedded at the bottom of the submitted manuscript. Kindly ensure that each table and figure is mentioned in the text and in numerical order. Figures must be of sufficient resolution for publication \href{http://home.frontiersin.org/about/author-guidelines#ResolutionRequirements}{see here for examples and minimum requirements}. Figures which are not according to the guidelines will cause substantial delay during the production process. Please see \href{http://home.frontiersin.org/about/author-guidelines#GeneralStyleGuidelinesforFigures}{here} for full figure guidelines. Cite figures with subfigures as figure \ref{fig:2}B.

\subsubsection{Permission to Reuse and Copyright}
Figures, tables, and images will be published under a Creative Commons CC-BY licence and permission must be obtained for use of copyrighted material from other sources (including re-published/adapted/modified/partial figures and images from the internet). It is the responsibility of the authors to acquire the licenses, to follow any citation instructions requested by third-party rights holders, and cover any supplementary charges.
%%Figures, tables, and images will be published under a Creative Commons CC-BY licence and permission must be obtained for use of copyrighted material from other sources (including re-published/adapted/modified/partial figures and images from the internet). It is the responsibility of the authors to acquire the licenses, to follow any citation instructions requested by third-party rights holders, and cover any supplementary charges.

\subsection{Tables}
Tables should be inserted at the end of the manuscript. Please build your table directly in LaTeX.Tables provided as jpeg/tiff files will not be accepted. Please note that very large tables (covering several pages) cannot be included in the final PDF for reasons of space. These tables will be published as \href{http://home.frontiersin.org/about/author-guidelines#SupplementaryMaterial}{Supplementary Material} on the online article page at the time of acceptance. The author will be notified during the typesetting of the final article if this is the case.

@@

\subsection{Resource Identification Initiative}
To take part in the Resource Identification Initiative, please use the corresponding catalog number and RRID in your current manuscript. For more information about the project and for steps on how to search for an RRID, please click \href{http://www.frontiersin.org/files/pdf/letter_to_author.pdf}{here}.

\subsection{Life Science Identifiers}
Life Science Identifiers (LSIDs) for ZOOBANK registered names or nomenclatural acts should be listed in the manuscript before the keywords. For more information on LSIDs please see \href{http://www.frontiersin.org/about/AuthorGuidelines#InclusionofZoologicalNomenclature}{Inclusion of Zoological Nomenclature} section of the guidelines.

\section{Additional Requirements}

For additional requirements for specific article types and further information please refer to \href{http://www.frontiersin.org/about/AuthorGuidelines#AdditionalRequirements}{Author Guidelines}.

\section*{Conflict of Interest Statement}
%All financial, commercial or other relationships that might be perceived by the academic community as representing a potential conflict of interest must be disclosed. If no such relationship exists, authors will be asked to confirm the following statement:

@@

The Author Contributions section is mandatory for all articles, including articles by sole authors. If an appropriate statement is not provided on submission, a standard one will be inserted during the production process. The Author Contributions statement must describe the contributions of individual authors referred to by their initials and, in doing so, all authors agree to be accountable for the content of the work. Please see  \href{http://home.frontiersin.org/about/author-guidelines#AuthorandContributors}{here} for full authorship criteria.

\section*{Funding}
Details of all funding sources should be provided, including grant numbers if applicable. Please ensure to add all necessary funding information, as after publication this is no longer possible.

\section*{Acknowledgments}
This is a short text to acknowledge the contributions of specific colleagues, institutions, or agencies that aided the efforts of the authors.

\section*{Supplemental Data}
 \href{http://home.frontiersin.org/about/author-guidelines#SupplementaryMaterial}{Supplementary Material} should be uploaded separately on submission, if there are Supplementary Figures, please include the caption in the same file as the figure. LaTeX Supplementary Material templates can be found in the Frontiers LaTeX folder.

\section*{Data Availability Statement}
The datasets [GENERATED/ANALYZED] for this study can be found in the [NAME OF REPOSITORY] [LINK].
% Please see the availability of data guidelines for more information, at https://www.frontiersin.org/about/author-guidelines#AvailabilityofData

@@

%%% Make sure to upload the bib file along with the tex file and PDF
%%% Please see the test.bib file for some examples of references

\section*{Figure captions}

%%% Please be aware that for original research articles we only permit a combined number of 15 figures and tables, one figure with multiple subfigures will count as only one figure.
%%% Use this if adding the figures directly in the mansucript, if so, please remember to also upload the files when submitting your article
%%% There is no need for adding the file termination, as long as you indicate where the file is saved. In the examples below the files (logo1.eps and logos.eps) are in the Frontiers LaTeX folder
%%% If using *.tif files convert them to .jpg or .png
%%%  NB logo1.eps is required in the path in order to correctly compile front page header %%%

\begin{figure}[h!]
\begin{center}
\includegraphics[width=10cm]{logo1}% This is a *.eps file
\end{center}
\caption{ Enter the caption for your figure here.  Repeat as  necessary for each of your figures}\label{fig:1}
\end{figure}

\begin{figure}[h!]
\begin{center}
\includegraphics[width=15cm]{logos}
\end{center}
\caption{This is a figure with sub figures, \textbf{(A)} is one logo, \textbf{(B)} is a different logo.}\label{fig:2}
\end{figure}

%%% If you are submitting a figure with subfigures please combine these into one image file with part labels integrated.
%%% If you don't add the figures in the LaTeX files, please upload them when submitting the article.
%%% Frontiers will add the figures at the end of the provisional pdf automatically
%%% The use of LaTeX coding to draw Diagrams/Figures/Structures should be avoided. They should be external callouts including graphics.
